### 2020-02-01 paper/sex_selection_spacing-ver3.tex
Typo in package name
Need complete new abstract and various edits

\usepackage{endfloat}`

@@

% Demography abstract
\noindent 
Strong son preference is typically associated with shorter birth spacing in the absence of 
sons, but access to sex selection has the potential to reverse this pattern because each 
abortion extends spacing by six to twelve months. 
I apply an empirical method that simultaneously accounts for how sex selection 
increases both birth spacing and the likelihood of a son. 
Using the four rounds of India's National Family and Health Surveys, I show that birth 
intervals increased substantially over the last four decades, with the most significant
increases among women most likely to use sex selection.
Specifically, well-educated women with no boys now have significantly longer birth
intervals, and more male-biased sex ratios, than similar women with boys. 
Women with no education still follow the standard pattern of short average spacing when 
they have girls and little evidence of sex selection, with medium-educated women showing 
mixed results.
Furthermore, the traditional pattern of longer average spacing with sons appears to arise 
not from fewer very short birth intervals, but rather from the upper end of the distribution.

@@

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhargava2003,Bhalotra2008,Maitra2008,
Jayachandran2011,Jayachandran2017a}.
Longer spacing between births then potentially lowers mortality, although there may
be counteracting effects if the longer spacing arises because of multiple abortions.
There has so far been no attempt to examine which mechanism dominates.

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals, the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Now, longer birth intervals in the absence of sons may instead arise because couples 
with a strong preference for sons use sex selection, possibly to the point where we 
observe a reversal of the traditional spacing pattern.
Despite a large literature on son preference and birth spacing, there is, however,
no prior research that has examined how sex selection affects spacing.

@@

First, how did average birth intervals change over time for different groups, and how do 
these changes relate to the use of sex selection?
To show how sex selection has affected birth spacing, I focus on how birth intervals have
changed for different education levels, urban or rural, and the sex composition of 
prior children.

@@

Second, how has the distribution of birth spacing changed?
Average birth intervals help us understand the overall changes in 
spacing behavior, but tells us little about how these changes came about.
Examining how the distribution has change allow us to separate out increases in birth
spacing that come from women with strong son preference who still try to conceive early, 
but are also the most likely to end up with very long spells because of multiple 
abortions versus women who have longer average spacing because they try to avoid very
short birth intervals.

@@

Finally, has the relationship between birth spacing and mortality changed over time?
[TK]

[TK results]

@@

To set the stage for the subsequent analyses, I first describe how female education
and labor force participation has changed.
Both are potentially important factors in the birth spacing decision, as I discuss
in detail in the subsequent section.

@@

The substantial increases in spacing for women with only girls is not evidence of lower
son preference as shown by the changes in the predicted sex ratios and the parity 
progression behavior.
For both the third and fourth spell the predicted sex ratios in urban and rural areas are
statistically significantly above the natural sex ratio at between 55.5\% and 57.7\% boys.
The elevated sex ratios strongly suggest that sex selection, possibly precipitated by
declining fertility, drove the increases in birth intervals for women with only girls and 
that without sex selection we would instead have observed a diverging pattern in birth 
intervals across sex composition similar to what we saw for women without education.

@@

Women with eight or more years of education are expected to have the lowest fertility and 
the highest use of sex selection, and we should, therefore, see the most substantial 
effects on spacing.
At first blush, it appears from Figure \ref{fig:bootstrap_high} that the second spell 
spacing pattern looks remarkably similar to the two other education groups with little 
difference in average birth spacing across sex composition.
This similarity hides, however, a much faster decline in the likelihood of having a second 
birth, substantial increases in spacing, and predicted sex ratios for women with a 
first-born daughter above 55 percent for both urban and rural areas for the last two 
periods.
Hence, where it not for a substantial amount of sex selection brought on by falling
fertility we would now have seen a standard son preference spacing pattern with 
significantly shorter spacing after a first-born girl than after a first-born boy.

>>>>

\usepackage{endfloat}

@@

% Demography abstract
\noindent

@@

Second, longer birth spacing is associated with lower mortality and 
morbidity risk \citep{Conde-Agudelo2012,Molitoris2019}.
In India, the duration to the next birth has traditionally been shorter with fewer sons, 
which likely contributed to the higher mortality and morbidity risk for girls
\citep{Whitworth2002,Bhargava2003,Bhalotra2008,Maitra2008,
Jayachandran2011,Jayachandran2017a}.
Longer spacing between births then potentially lowers mortality, although there may
be counteracting effects if the longer spacing arises because of multiple abortions, but
there has so far been no attempt to examine which mechanism dominates.

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals, the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Now, longer birth intervals in the absence of sons may instead arise because couples 
with a strong preference for sons use sex selection, possibly to the point where we 
observe a reversal of the traditional spacing pattern.
Despite a large literature on son preference and birth spacing, there is, however,
no prior research has examined how sex selection affects spacing.

@@

First, how did average birth intervals change over time for different groups, and how do 
these changes relate to the use of sex selection?

@@

[Shorten/combine these two paragraphs]
Second, how has the distribution of birth spacing changed?
Average birth intervals help us understand the overall changes in 
spacing behavior, but tells us little about how these changes came about.
Examining how the distribution has change allow us to separate out increases in birth
spacing that come from women with strong son preference who still try to conceive early, 
but are also the most likely to end up with very long spells because of multiple 
abortions versus women who have longer average spacing because they try to avoid very
short birth intervals.

@@

Finally, how has the relationship between birth spacing 
and mortality changed over time?
Mortality risk is inversely related to education level,
especially for very short birth intervals.
However, infant mortality risk has decline substantial 
over time for the lower education groups and is now
close to the level of women with the most education.
There is no evidence that the increasing use of 
sex selection is associated with higher mortality risk.

@@

To set the stage for the subsequent analyses, I first describe how female education
and labor force participation has changed.
Both are potentially important factors in the birth spacing decision, as I discuss
in the subsequent section.

@@

The substantial increases in spacing for women with only girls is not evidence of lower
son preference as shown by the changes in the predicted sex ratios and the parity 
progression behavior.
For both the third and fourth spell the predicted sex ratios in urban and rural areas are
statistically significantly above the natural sex ratio at between 55.5\% and 57.7\% boys.
The elevated sex ratios strongly suggest that sex selection, possibly precipitated by
declining desired fertility, drove the increases in birth intervals for women with only girls and 
that without sex selection we would instead have observed a diverging pattern in birth 
intervals across sex composition similar to what we saw for women without education.

@@

Women with eight to eleven years of education are 
expected to have lower fertility and 
higher use of sex selection, and we should, therefore, 
see substantial effects on spacing.
At first blush, it appears from Figure \ref{fig:bootstrap_high} that the second spell 
spacing pattern looks remarkably similar to the two other education groups with little 
difference in average birth spacing across sex composition.
This similarity hides, however, a much faster decline in the likelihood of having a second 
birth, substantial increases in spacing, and predicted sex ratios for women with a 
first-born daughter above 55 percent for both urban and rural areas for the last two 
periods.
Hence, where it not for a substantial amount of sex selection brought on by falling
fertility we would now have seen a standard son preference spacing pattern with 
significantly shorter spacing after a first-born girl than after a first-born boy.
